Choose a License

### 2025-01-25 choose_license.qmd
Major rewrite
First explain copyright and licenses,
then practice citation and attribution,
then choose own license.

Create file `copyright.qmd` with material
from `choose_license.qmd` and with
additional information on license conditions
and providing attribution.

Create files `data.qmd` and `code.qmd` with
material from `setup.qmd`. Immediately
cover the citation and attribution.

Relicense `Manuscript.qmd` and all code chunks under CC0 1.0. Relicense other
materials under CC BY-SA 4.0.

## TL;DR

Every expression of an idea, such as a literary or an artistic works, is automatically covered by copyright. Put simply, this means that nobody else than the copyright holder is allowed to copy, modify, or share it. Copyright licenses can grant others some of the necessary rights, but mostly don't cover publicity, privacy, moral, patent, or trademark rights. Certain exceptions and limitations to copyright (such as fair use or the right to quote) may obviate the need for a license.

Researchers are only allowed to share material they have the necessary rights to. This includes copyrights, which are granted by free/open copyright licenses. These licenses often require to attribute the authors, indicate whether changes were made, and provide the text of the license, among other things. In addition, researchers should share any work by their own under such a license.

_Choosing a license_ involves:

1. Recording the copyright notices and licenses for content by others (for example, in a file called `LICENSE.txt`) and following all their additional requirements.
2. In the simplest case, choosing two licenses (one suitable for code, one for everything else) which to apply  simultaneously to all own content (for example, again by recording the license in the file `LICENSE.txt`). Recipients are then free to choose under which of the two licenses to use the material.

We recommend to choose [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) and [Apache\ 2.0](https://choosealicense.com/licenses/apache-2.0/) for all own works. Take your time to read the summaries behind the links and/or the actual license text to understand its legal effect.

You can adapt the following wording to your use case:

> Except where noted otherwise, all files in this project are made available under CC0\ 1.0 or (at your option) under the terms of the Apache Software License\ 2.0.

## A Primer on Licenses

Whenever you create a literary or artistic work (such as a text, image, video, or software), the copyright law in most countries limits other people from copying, modifying, and sharing it without your express permission. [If the work was created as part of your job, it might be your employer who holds the copyright to the work, depending on the country and contract.]{.aside} This even applies if you make it available to others (e.g., on your website): First and foremost they are not allowed to copy, modify, or share it. This legal default of "all rights reserved" was created to benefit publishers, not authors [@Fogel2006], and runs counter to many cultural and scientific processes. Copyright licenses enable authors to free up their works for reuse by others.

::: {#nte-situation-lmu .callout-note collapse="true"}
### For researchers at LMU

A license is a legal document that regulates what _others_ are allowed to do with a copyright-protected work -- the licenses we discuss are non-exclusive licenses and do not limit the author or copyright holder in their rights. And while you _could_ write your own license, there are already many pre-formulated licenses available to choose from and to apply to your work.^[In fact, you should not create your own license if the purpose is to share your work freely and openly with others. By using a boilerplate license others know what to expect, also because they have been [tested](https://sfconservancy.org/copyleft-compliance/) in [courts](https://legaldb.creativecommons.org/).]

::: {.column-margin}
For the purpose of this tutorial, by _license_ we mean _copyright license_.
:::

@@

The licenses we discuss here mostly regard copyright. Therefore, recipients may lack other rights such as publicity, privacy, moral, patent,^[In fact, the software licenses we recommend on this page have been specifically selected to provide an express grant of patent rights.] or trademark rights. For example, sharing photos that depict people is not only a matter of copyright, but also of privacy rights. Conversely, the licenses do not apply if recipients are allowed to use the works for other reasons such as fair use, the right to quote, or because they made a different arrangement with the author.

@@

Many boilerplate licenses are available to apply to your work. Which license is appropriate depends on several factors, including existing licenses in place and the type of work, but also your personal considerations. We strongly recommend to apply a _free/open_ license to your work, which means that the work "can be freely studied, applied, copied and/or modified, by anyone, for any purpose" [@Moller2015]. Importantly, this also means that others do not need to ask or notify the author and that they can use it for commercial purposes. By the choice of license, authors can, however, demand that they are credited, that the original license is indicated, that modifications are indicated, that derivative works are only shared under the same license, and that no further restrictions are imposed on the work. Software licenses may additionally require to make the source code available to everybody the software is shared with and often require to display the full text of the license upon usage. Because there are many free/open licenses available, the licenses discussed here only represent a recommended subset.

::: {.column-margin}
If you would like to choose a license not listed here, it should be appropriate for the type of work in question and be compatible with the dominant copyleft license in the respective community (see also @nte-silos). For software that's almost universally the (A)GPLv3 and for writing, image, audio, and video that's mostly the [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/). For data, no dominant copyleft license has emerged yet, so any of [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/), [CDLA\ Sharing\ 1.0](https://cdla.dev/sharing-1-0/), and [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) are acceptable.
:::

::: {#nte-terminology .callout-note collapse="true"}
### A Note on Terminology

The terms _free_ and _open_, especially with regard to software, come with a history. The Free Software Foundation (FSF) was founded in 1985 to protect "four essential freedoms" [-@FSF2024] of a program's user. These are the freedoms to use, study, share, and improve a program. Software whose users legally and practically have these freedoms (because, among other things, they have access to its source code) is considered _free_. The four freedoms are seen as vital for a society as a whole in the sense that they enable sharing, cooperation and ultimately freedom in general. The FSF maintains [a list of software licenses](https://www.gnu.org/licenses/license-list.en.html#SoftwareLicenses) that it considers to be protecting the four freedoms. Sometimes the term _libre_ (Spanish and French for _free_) is used to make a distinction from _gratis_ software. You can learn more about free software at [Write Free Software](https://writefreesoftware.org/learn).

The Open Source Initiative (OSI), which was founded in 1998, follows a more pragmatic approach. It is concerned with developing high-quality software, for which everyone's ability to obtain, modify and contribute back the source code is considered beneficial. Access to the source code is one out of multiple conditions for software to be considered _open source_ by the OSI [-@OSI2007], which equally maintains [a list of approved licenses](https://opensource.org/licenses).

Access to the source code is a necessary, but not sufficient requirement both for _free software_ and for _open source software_. Conversely, both can (and frequently are) sold for money, as their respective criteria only apply once one has access to the software. Throughout this tutorial, we write "free/open license" to mean a license that is approved by either the FSF or the OSI. Software which is neither but makes available its source code is sometimes referred to as _source-available software_.

Apart from software, the term _open access_ has often been used for works that are available at no cost. For example, this is the commonality of bronze, green, hybrid, gold, and diamond/platinum open access _articles_, which otherwise vary in the rights that are granted to readers. In 2002, the Budapest Open Access Initiative declared that _open access_ additionally includes the right to use articles for any purpose, and in 2003, the Bethesda Statement and the Berlin Declaration added the right to make derivative works.

Two other notable definitions include the _Open Definition_ [@OKFN2016], which was first drafted in 2005, and the definition of _Free Cultural Works_ [@Moller2015], for which the open editing phase began in 2006. They are largely viewed as compatible with one another.

:::: {.column-margin}
![](images/Definition_of_Free_Cultural_Works.svg){width=250px}
::::
:::

@@

First, if you adapt (i.e., modify, remix, build on) a work by others you need to determine if it is provided to you under a free/open license.[You can determine whether a license is free/open by searching for its name in the [SPDX License List](https://spdx.org/licenses/) and looking for at least one `Y` in the two columns _FSF Free/Libre?_ and _OSI Approved?_]{.aside} If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] or if a particular license choice is mandated by your funder. Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

If you create a new work and no strong community norms suggest a particular license, you need to choose the license yourself. Which license to choose depends on the type of work you create. Software licenses, for example, may consider that the source code is the preferred form for making modifications, while licenses for data can differentiate between the database and any works produced from it. We have created a flowchart that covers the most likely types of works you will create as a researcher: software, writing (i.e., text), images, audio, video, and data (see @fig-flowchart-simple). This flowchart always recommends the most permissive license possible to maximize reuse -- though we provide two additional flowcharts below that allow for more choices. Click on the name of a license to learn more about it.

@@

One may wonder which license to apply in this case, because Creative Commons licenses are not recommended for source code^[because (among other reasons) they explicitly disclaim any conveyance of patent rights] and applying software licenses to PDFs or images can lead to confusion or nuisance.^[because they often require to display the full text of the license]

@@

> The Quarto files in this project are made available under CC0\ 1.0 or (at your option) under the terms of the Apache Software License\ 2.0.
:::

Note, that for data there is, at least in principle, the possibility that a different license applies to the individual entries than to the collective database. For example, if you were to create a database of artworks by others, those artworks would be licensed individually as chosen by the artists, but the license for the database as a whole could be chosen by you. The latter includes the selection of entries, field names and any _sui generis_ database rights. However, if the content was created by you, we recommend you to choose the same license for both content and database. Factual data (like measurements or metadata) should be licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) [@Villa2016Copyleft; @Villa2016CC0] -- otherwise, consult @nte-other-work-types for some caveats.

@@

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. The [SIL Open Font License\ 1.1](https://openfontlicense.org/) is the _de facto_ standard license for fonts, requiring that derived fonts, if published, have to be put under the same license. It doesn't require attribution for usage (e.g., when embedded in a document), but forbids that the font is sold by itself. Selling the font is allowed, however, if it is bundled with other software.
- __templates and TeX packages:__ If a template or, for this purpose, a $\TeX$ package, is licensed under a copyleft license such as the (A)GPLv3, every work that is a derivative of the template or that uses the $\TeX$ package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the (A)GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to be licensed under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or another license).
:::

@@

The consequence of applying a permissive license to your work is that others may distribute their adaptations from it under stricter terms, possibly without a free/open license. You may prefer that everybody who receives somebody else's adaptations of your work enjoys the same rights that you granted to them. This can be achieved by _copyleft_ licenses.^[So called because copyleft licenses make content permanently free, thus turning copyright around: © → 🄯. In the realm of Creative Commons licenses, copyleft is also called _share alike_.] For software, they come in two flavors:

- __Weak copyleft licenses__ only require that modifications to the software itself are licensed under the same or a compatible license if shared. For example, if you create and publish software under a weak copyleft license, others who modify it and put their version on the internet have to apply the same license. However, people who merely _use_ your software in their own work which they make publicly available can choose any license for it.
- __Strong copyleft licenses,__ on the other side, insist that any larger works that use the copyleft-licensed software must also be licensed under the same or a compatible license if shared. For example, if your software were to be put under a strong copyleft license, everybody publishing software that uses your software would need to put it under the same license. Because of only few rulings in courts, the extent of this requirement is disputed [@Wikipedia2024].

Note, however, that the copyleft licenses we discuss here do not mandate sharing. Copyleft (and attribution) clauses are only triggered if the work is shared [@CC2015]. This means that if you only use a work internally, you do not need to share your derivative works. It is also worth reiterating that these licenses do not restrict the original author(s): They are still permitted to distribute their work under a different license and without sharing the source code.

@@

In most cases, the output of software, like images or tables, does not depend on the software's license.^[The exception is when the software is part of the output [see @apsillers2019].] Therefore, if you use an R package under a copyleft license to create a figure, you are likely the copyright owner. However, if the output is based on data, it can be considered a derivative work of the data and the license of the data also applies. For example, maps may be considered as a derivative of the geographic data they are based on.

@@

```r
deps <- renv::dependencies()$Package |>
  unique() |>
  pak::pkg_deps(dependencies = NA) |>
  getElement("package")
unique(installed.packages(fields="License")[deps, "License"])
```
:::

@@

### Attribution, State Changes, and Anti-DRM?

The advanced license flowcharts also allow you to make additional decisions:

- __Attribution__ means whether recipients of your work are required to provide attribution to you. For software licenses, the attribution is part of a copyright notice. Note that even if attribution is not a requirement of the license, good scientific practice demands that appropriate citations are made.
- __Add license text__ means that recipients need to distribute the full text of the license along with the work (rather than only mentioning its license, for example, via a link).
- __State changes__ means that recipients need to indicate if changes were made.
- __Disclose code__ means that recipients need to disclose the source code when distributing modified versions of your work.
- __Anti-DRM__ means that when others share your work they are not allowed to apply technological measures restricting anything that the license permits. Implementing such restrictive measures is also referred to as **d**igital **r**ights **m**anagement (DRM).

From the Creative Commons licenses, only [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) does not require providing attribution. All software licenses in @fig-flowchart-non-data require providing attribution, although only [Apache\ 2.0](https://choosealicense.com/licenses/apache-2.0/) and [AGPLv3](https://choosealicense.com/licenses/agpl-3.0/) require others to indicate changes. For data, the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) allows for technological measures that restrict the work only if a version of the database is provided in parallel without such measures.

@@

As indicated before, a free/open license must allow creating derivative works and must allow exercising the rights granted by it _for any purpose_, including commercial use. Otherwise, creating adaptations from multiple works by different authors would either limit commercial use for everybody (including the authors) or would not be possible at all (when including materials with a copyleft license), effectively creating _silos_ of works which are mutually incompatible with each other. This is especially relevant for work types for which adaptations are created frequently, such as text, data, and source code. In order to avoid silos, one should only choose licenses which are compatible with the dominant copyleft license in the respective community [@Lammerhirt2017; @Wheeler2014]. If you would like to learn more about the different types of compatibility, we recommend you to read the article "A Quick Guide to Software Licensing for the Scientist-Programmer" by @Morin2012. The following diagram provides an overview of the compatibility of various licenses:

@@

## Which Licenses to Choose for a Project?

So far, we only discussed how to choose a license for works _of one type_. But what if you want to share a project with all kinds of files? For example, the project from this tutorial (among other things) includes a data file, a manuscript file with intermingled code and writing, and an R file. And what if you also want to share files by others, as is the case with `apaquarto` which you may have installed in this project?

The answer is that you need to indicate the license on a per-file or per-folder basis (rather than choosing one for the whole project). Sometimes, even different parts of a file might be subject to different licenses. The easiest approach is to make note of every foreign work included in your project and record its license. Then, multi-license all the remaining files, which are yours, under a code and a non-code license in parallel as explained in @tip-license-r.

@@

Having selected the licenses of your choice -- again, you might need multiple ones depending on the types of works your project contains --, we encourage you to read through the full license text (or at least a legal summary) to understand their effect. Then, you can record the license of existing content and apply the licenses of your own contributions. Mostly, this just means indicating which file or folder is covered by which license, usually in the project's README (among other places), whose creation will be discussed [later](make_readme.qmd). Notably, no sort of registration is required.

::: {#cau-license-versions .callout-caution}
### License Versions Are Important

You may have noticed that we mostly refer to licenses using a name _and_ a version number. This is because the organizations that created the licenses sometimes publish updated versions to accommodate for developments in copyright law and the communities that use the licenses. For example, the Creative Commons licenses (that start with `CC`) were first published in 2002. Since then, the possibility to relicense under later ([v2.0](https://creativecommons.org/weblog/entry/4216#:~:text=Share%20Alike%20Across%20Borders), 2004) and compatible licenses ([v3.0](https://creativecommons.org/weblog/entry/7249), 2007) has been added, a 30-day window to correct license violations has been established to combat [copyleft trolls](https://commons.wikimedia.org/wiki/Commons:Copyleft_trolling), and _sui generis_ database rights are covered explicitly ([v4.0](https://creativecommons.org/weblog/entry/40768), 2013). There are many more [subtle differences between license versions](https://wiki.creativecommons.org/wiki/License_Versions), therefore it is important to indicate which license version exactly one is referring to, as the license of a work does not "update" automatically. CC0, officially introduced in 2009, is still at version [1.0](http://www.creativecommons.org/weblog/entry/13304).

For the AGPLv3 it is even recommended to state whether a work is licensed under exactly the indicated version of the license or, alternatively, also under newer versions of the license [@Stallman2022Version].
:::

For example, the `apaquarto` extension that you included in your project is a work by others.^[Unless you are the author of the `apaquarto` extension.] You need to indicate its license so that others know what they are allowed to do -- and, of course, you need to comply with any terms yourself, such as retaining the copyright notice.^[In this particular case, `apaquarto` is licensed under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/), which does not require you to retain a copyright notice. In fact, it's an extremely permissive license.] In contrast, if it were not for this tutorial, the manuscript would contain your own work and you would need to indicate under which license you provide it to others.

@@

While it is common to state the chosen license(s) in the README, usually one of the following actions is taken in addition:

In the simplest case, one just creates a file called `LICENSE.txt` where the full text of the license is copied verbatim. This is a practice propagated by GitHub, which provides instructions for and comparisons of many licenses via [ChooseALicense.com](https://choosealicense.com/licenses/). However, if the project is not completely covered by one single license, this practice may become unwieldy. For example, if a project contains different types of works by different authors, the `LICENSE.txt` needs to detail which file is covered by which license(s), along with any copyright notices.

Individual programming languages also have their own way of stating which license a package is distributed under. For R packages, this is usually set by the field `License` in the file `DESCRIPTION` [@Wickham2023PackagesLicense].

Finally, @nte-reuse explains how to use the [REUSE](https://reuse.software/) specification to make the choice of license machine-readable. This is the approach we recommend taking.

@@

Every major free/open license has a unique [SPDX identifier](https://spdx.org/licenses/) which allows communicating the license choice unequivocally. We will be using that to indicate the license for every file in your project, along with the year of publication and the copyright holder. To do this, we add a comment to the beginning of every file and include the two tags `SPDX-FileCopyrightText` and `SPDX-License-Identifier`. How this works depends on the file type, as the syntax for a comment varies.

@@

Regardless of how exactly the licenses are added to the project, this is a good opportunity to verify one last time that all third party content is provided to you under a free/open license and that you comply with it. Please add a license to your project now, either creating a file `LICENSE.txt` or following the REUSE standard.

## Wrap-up

If you would like to learn more about copyright and licenses you might find the following resources interesting:

- "Open Content -- A Practical Guide to Using Creative Commons Licences" by @Kreutzer2014
- "Creative Commons Certificate for Educators, Academic Librarians, and Open Culture" by @CC2024Certificate
- "Freie Software -- Zwischen Privat- und Gemeineigentum" by @Grassmuck2004

>>>>

So far, you took care to legally include works by others in your project folder. Up next, you will free up your project for reuse. This means ensuring that every bit (co-)authored by you in the project is put under a free/open license. It may be that your project (i.e., its files and folders) is not one work, but rather consists of multiple works available under different licenses. In that case, you need to indicate the license on a per-file or per-folder basis (rather than choosing one for the whole project). Sometimes, even different parts of a file might be subject to different licenses. For the purpose of this tutorial, we will consider the manuscript which you edited and the data dictionary you created.

@@

::: {#nte-license-choice-tldr .callout-note}
## License Choice TL;DR

We recommend to choose free/open licenses suitable to the type of work (i.e., code, data, and everything else). [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) and [Apache\ 2.0](https://choosealicense.com/licenses/apache-2.0/) are good choices to apply simultaneously to all own works. Take your time to read the summaries behind the links and/or the actual license text to understand the legal effect. You can adapt the following wording to your use case:

> Except where noted otherwise, all files in this project are made available under CC0\ 1.0 or (at your option) under the terms of the Apache License\ 2.0.
:::

@@

Many boilerplate licenses are available to apply to your work. Which license is appropriate depends on several factors, including existing licenses in place and the type of work, but also your personal considerations. We strongly recommend to apply a _free/open_ license to your work that fits to the work type and potential use cases. Because there are many free/open licenses available, the licenses discussed here only represent a recommended subset.

::: {.column-margin}
If you would like to choose a license not listed here, it should be appropriate for the type of work in question and be compatible with the dominant copyleft license in the respective community (see also @nte-silos). For software that's almost universally the (A)GPLv3 and for writing, image, audio, and video that's mostly the [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/). For databases, no dominant copyleft license has emerged yet, so any of [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/), [CDLA\ Sharing\ 1.0](https://cdla.dev/sharing-1-0/), and [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) are acceptable.
:::

@@

First, if you adapt (i.e., modify, remix, build on) a work by others you need to determine if it is provided to you under a free/open license. If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] or if a particular license choice is mandated by your funder. Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

If you create a new work and no strong community norms suggest a particular license, you need to choose the license yourself. Which license to choose depends on the type of work you create. We have created a flowchart that covers the most likely types of works you will create as a researcher: software, writing (i.e., text), images, audio, video, and data (see @fig-flowchart-simple). This flowchart always recommends the most permissive license possible to maximize reuse -- though we provide two additional flowcharts below that allow for more choices. Click on the name of a license to learn more about it.

@@

One may wonder which license to apply in this case, because Creative Commons licenses are not recommended for source code^[because (among other reasons) they explicitly disclaim any conveyance of patent rights [see @CC2024Software]] and applying software licenses to PDFs or images can lead to confusion or nuisance.^[because they often require to display the full text of the license [see @Wikipedia2024FDL]]

@@

> The Quarto documents in this project are made available under CC0\ 1.0 or (at your option) under the terms of the Apache License\ 2.0.
:::

We already mentioned that a database may be licensed differently than its content. However, if the content was created by you, we recommend you to choose the same license for both content and database. Factual data (like measurements or metadata) should be made available with [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) [@Villa2016Copyleft; @Villa2016CC0] -- otherwise, consult @nte-other-work-types for some caveats.

@@

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. The [SIL Open Font License\ 1.1](https://openfontlicense.org/) is the _de facto_ standard license for fonts, requiring that derived fonts, if published, have to be put under the same license. It doesn't require attribution for usage (e.g., when embedded in a document), but forbids that the font is sold by itself. Selling the font is allowed, however, if it is bundled with other software.
- __templates and TeX packages:__ If a template or, for this purpose, a $\TeX$ package, is licensed under a copyleft license such as the (A)GPLv3, every work that is a derivative of it has to be put under the same license if shared. If a document contained source code covered by the (A)GPLv3, it depends on the individual case whether the same license needs to be applied to the document if shared.
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to be licensed under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or another license).
:::

@@

You may prefer that adaptations of your work stay free/open and thus choose a copyleft license. Of course, this does not restrict you as the original author: You are still permitted to distribute the work under a different license and without sharing the source code.

@@

In most cases, the output of software, like images or tables, does not depend on the software's license.^[The exception is when the software is part of the output [see @FSF2024].] Therefore, if you use an R package under a copyleft license to create a figure, you are likely the copyright owner. However, if the output is based on data, it can be considered a derivative work of the data and the license of the data also applies. For example, maps may be considered as a derivative of the geographic data they are based on.

@@

```{.r filename="Console"}
deps <- renv::dependencies()$Package |>
  unique() |>
  pak::pkg_deps(dependencies = NA) |>
  getElement("package")
unique(installed.packages(fields="License")[deps, "License"])
```
:::

@@

### Additional considerations

The advanced license flowcharts also allow you to make additional decisions, such as whether attribution should be a condition of redistribution. As you have learned, even if attribution is not a condition of the license, there are other reasons for attributing you as an author (such as scientific citation standards or moral rights). All conditions are explained on the page on [copyright](copyright.qmd).

From the Creative Commons licenses, only [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) does not require providing attribution, and all software licenses in @fig-flowchart-non-data require providing attribution. As an author, you are free to waive license conditions such as attribution, however.

@@

Creative Commons also offers licenses which prevent commercial use of a work and/or sharing of derivative works. A free/open license, however, must allow creating derivative works and must allow exercising the rights granted by it _for any purpose_, including commercial use. Otherwise, creating adaptations from multiple works by different authors would either limit commercial use for everybody (including the authors) or would not be possible at all (when including materials with a copyleft license), effectively creating _silos_ of works which are mutually incompatible with each other. This is especially relevant for work types for which adaptations are created frequently, such as text, data, and source code. In order to avoid silos, one should only choose licenses which are compatible with the dominant copyleft license in the respective community [@Lammerhirt2017; @Wheeler2014]. If you would like to learn more about the different types of compatibility, we recommend you to read the article "A Quick Guide to Software Licensing for the Scientist-Programmer" by @Morin2012. The following diagram provides an overview of the compatibility of various licenses:

@@

@@

Having selected the license(s) of your choice, we encourage you to read through the full license text (or at least a legal summary) to understand the legal effect. The following applies to all licenses mentioned here:

- Applying the license is irrevocable. You may stop distributing your work, but everybody who already received it may continue to share it.
- As an author applying the license, you are not bound by the license terms. You cannot, however, give others an exclusive license to your content anymore (as some journals needlessly demand, see @wrn-exclusive-rights).
- By applying the license you share with others your right to use your work for commercial purposes, including selling it.
- If you apply the license to an image, the license [may also apply](https://commons.wikimedia.org/wiki/Commons:Same_work) to higher resolution version of your image.
- If you apply a copyleft license to an image, a document in which the image is used may not constitute a derivative work and, as such, may not be affected by the copyleft condition.
- If you are a member of a collecting society^[in Germany: _Verwertungsgesellschaft_] or if the work was created as part of your job, you may not be able to apply the license. Check with your collecting society or your employer in that case.

::: {#nte-situation-lmu .callout-note collapse="true"}
### For Researchers at LMU Munich

For Creative Commons licenses, the following is additionally important to know:

- By applying the license you waive any moral, publicity, privacy, or other similar personality rights to the extent necessary to use the work.
- The attribution requirement may be implemented "in any reasonable manner", possibly not conforming to your precise expectations.

Applying a license to a work mostly just means indicating which file or folder is covered by which license(s) (no sort of registration is required). There are multiple standards on how to do so:

- You have already learned about the file `LICENSE.txt`. It usually contains detailed information, such as full license texts. This is a practice propagated by GitHub, which provides instructions for and comparisons of many licenses via [ChooseALicense.com](https://choosealicense.com/licenses/).
- A summary of the licensing situation is often given in a `README` file, whose creation will be discussed [later](make_readme.qmd).
- If only parts of a file are covered by a certain license, then it might make sense to indicate this directly in the file.
- The license of embedded elements (e.g., images) is either given directly beneath to where they are embedded, at the end of the respective page, or collectively on an separate page.
- Individual programming languages also have their own way of stating which license a package is distributed under. For R packages, this is usually set by the field `License` in the file `DESCRIPTION` [@Wickham2023PackagesLicense].
- Software projects often indicate the license in every file, following the [REUSE](https://reuse.software/) specification to make the choice of license machine-readable (explained in @nte-reuse).
- The following two metadata standards for increasing citability also allow to indicate the license:
  - [CodeMeta](https://codemeta.github.io/): All information are stored in a file called `codemeta.json`, which can be created [on their website](https://codemeta.github.io/codemeta-generator/). Alternatively, one can use the R package [`codemetar`](https://docs.ropensci.org/codemetar/) to create this file.
  - [Citation File Format](https://citation-file-format.github.io/) (CFF): Here, everything is stored in a file called `CITATION.cff`. It is used by GitHub to directly show a recommended citation for a repository. The file can be created [on their website](https://citation-file-format.github.io/cff-initializer-javascript/). Alternatively, the R package [`cffr`](https://cran.r-project.org/package=cffr) may be used for that purpose.

@@

@@

Every major free/open license has a unique [SPDX identifier](https://spdx.org/licenses/) which allows communicating the license choice unequivocally. We will be using that to indicate the license for every file in your project folder, along with the year of publication and the copyright holder. To do this, we add a comment to the beginning of every file and include the two tags `SPDX-FileCopyrightText` and `SPDX-License-Identifier`. How this works depends on the file type, as the syntax for a comment varies.

@@

The manuscript you downloaded for this tutorial is made available to you under [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) or (at your option) under the [Apache License\ 2.0](https://www.apache.org/licenses/LICENSE-2.0), as is all code in this tutorial (e.g., for creating the data dictionary). Therefore, you can apply any license to your modified version (as long as you follow the terms of the existing license). Please apply the license of your choice now, using any of the methods mentioned above (except the `README`, which we will cover later).

::: {#tip-manuscript-licensetxt .callout-tip collapse="true"}
### Addition to `LICENSE.txt` (Solution)

In the following example, the files from this tutorial are used under the terms of CC0\ 1.0 and therefore, the requirements of the Apache license\ 2.0 need not be followed.

The edited manuscript files have been multi-licensed under [AGPLv3](https://www.gnu.org/licenses/agpl-3.0.html) (or later) or (at option of the reuser) under [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), while the data dictionary and its code are made available under CC0\ 1.0.

```{.txt .code-overflow-wrap filename="LICENSE.txt"}
The Quarto documents stored in "Manuscript.qmd", "Manuscript.tex", and "Manuscript.pdf" by Josephine Zerna, Christoph Scheffel, and <YOUR NAME> are available under AGPLv3 (or later) <https://www.gnu.org/licenses/agpl-3.0.html> or (at your option) under CC BY-SA 4.0 <https://creativecommons.org/licenses/by-sa/4.0/>.

The data dictionary (stored in "data_dictionary.html") and the code used to create it (stored in "create_data_dictionary.R") by Florian Kohrt and <YOUR NAME> are available under CC0 1.0 <https://creativecommons.org/publicdomain/zero/1.0/>.
```

As the AGPLv3 requires reusers adding the full license text, we recommend to also copy it to `LICENSE.txt` (if not already in there).

<details>
<summary>Full text of AGPLv3</summary>

```{.txt .code-overflow-wrap filename="LICENSE.txt"}
                    GNU AFFERO GENERAL PUBLIC LICENSE
                       Version 3, 19 November 2007

 Copyright (C) 2007 Free Software Foundation, Inc. <https://fsf.org/>
 Everyone is permitted to copy and distribute verbatim copies
 of this license document, but changing it is not allowed.

                            Preamble

  The GNU Affero General Public License is a free, copyleft license for
software and other kinds of works, specifically designed to ensure
cooperation with the community in the case of network server software.

  The licenses for most software and other practical works are designed
to take away your freedom to share and change the works.  By contrast,
our General Public Licenses are intended to guarantee your freedom to
share and change all versions of a program--to make sure it remains free
software for all its users.

  When we speak of free software, we are referring to freedom, not
price.  Our General Public Licenses are designed to make sure that you
have the freedom to distribute copies of free software (and charge for
them if you wish), that you receive source code or can get it if you
want it, that you can change the software or use pieces of it in new
free programs, and that you know you can do these things.

  Developers that use our General Public Licenses protect your rights
with two steps: (1) assert copyright on the software, and (2) offer
you this License which gives you legal permission to copy, distribute
and/or modify the software.

  A secondary benefit of defending all users' freedom is that
improvements made in alternate versions of the program, if they
receive widespread use, become available for other developers to
incorporate.  Many developers of free software are heartened and
encouraged by the resulting cooperation.  However, in the case of
software used on network servers, this result may fail to come about.
The GNU General Public License permits making a modified version and
letting the public access it on a server without ever releasing its
source code to the public.

  The GNU Affero General Public License is designed specifically to
ensure that, in such cases, the modified source code becomes available
to the community.  It requires the operator of a network server to
provide the source code of the modified version running there to the
users of that server.  Therefore, public use of a modified version, on
a publicly accessible server, gives the public access to the source
code of the modified version.
